Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2025-04-04 paper/paper.tex
arxiv flags

% -*- fill-column: 80; -*-
%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
%% \documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For final camera-ready submission, w/ required CCS and ACM Reference
\documentclass[acmsmall]{acmart}\settopmatter{}

@@

%% Keywords
%% comma separated list
\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission

@@

This paper begins in \cref{sec:type-theory-examples} by studying small
example programs from \theoryabbv to build intuition.  From there, in
\cref{sec:tt} we provide the syntax, typing, and equational theory of
\theoryname. In \cref{sec:applications} we demonstrate the
applicability of \theoryabbv for relating familiar grammar and
automata formalisms as well as building concrete parsers.  Then in
\cref{sec:semantics-and-metatheory}, we give a denotational semantics
that makes precise the connection between \theoryabbv syntax and
formal grammars. Finally in \Cref{sec:discussion} we discuss related
and future work.

@@

We present a selection of the non-linear type constructors in
\Cref{fig:non-linear-types} and provide the rest in
\ifappendix{\cref{fig:full-non-linear-types,fig:jdg-eq-nonlinear} of
    \cref{sec:syntax}}\else{the appendix}\fi.  First,
we include universe types $U$ of small non-linear types and $L$ of linear
types. These are defined as
universes ``ala Coquand'' in that we define judgments saying when
non-linear and linear types are \emph{small} and define the universes to internalize
precisely this judgment \cite{coquandPresheafModel,lmcs:7713}. The definition of
smallness is simply that all types are small as long as their sub-formulae are,
with the exception of the two universe types themselves. A formal
description of smallness is given in
\ifappendix{\cref{fig:small-nonlin-ty,fig:small-lin-ty} of
    \cref{sec:syntax}}\else{the appendix}\fi. These universe
types are needed so that we can define non-linear and linear types by recursion on natural
numbers.
Next, we include standard $\Sigma,\Pi$, empty, unit, Boolean, and natural number
types. More complex inductive types --- such as sum types, list types, and
$Fin~n$ --- can be defined in terms of these primitives, and we give their
encoding in \ifappendix{\cref{sec:syntax}}\else{the appendix}\fi.

@@

We give the rules for linear type
formation in \cref{fig:linear-formation} and the definition of linear terms in
\cref{fig:linear-terms}. The equational theory for these
types is straightforward $\beta\eta$ equivalence and is included in
\ifappendix{\cref{fig:jdgeq} of \cref{sec:denotational}}\else{the appendix}\fi.

@@

Next, we introduce the most complex and important linear type
constructors of our development, \emph{indexed inductive linear
types}. We encode these by adding a mechanism for constructing initial
algebras of strictly positive functorial type expressions, following
prior work on inductive types
\cite{nakov_quantitative_2022,altenkirch_indexed_2015}. The syntax is
given in \Cref{fig:iilt}. First, we add a non-linear type $\SPF\,X$ of
\emph{strictly positive functorial} linear type expressions indexed by
a non-linear type $X$. We think of the elements of this type as
syntactic descriptions of linear types that are parameterized by
$X$-many variables standing for linear types that are only used in
strictly positive positions. Accordingly, the $\SPF\,X$ type supports
an operation $\el$ that interprets it as such a type constructor, as
well as an operator $\map$ that defines a functorial action on parse
transformers. The $\SPF\,X$ type supports constructors for a reference
$\Var~x$ to one of the linear type variables, a constant expression
that doesn't mention any type variables $K$, as well as tensor
products and additive conjunction and disjunction of type expressions.
Further, we add equations in \ifappendix{\cref{fig:spf-act} of \cref{sec:syntax}
  }\else{the appendix }\fi that say that the
$\el$/$\map$ operations correspond to these descriptions of the
constructors.

@@

\begin{figure}
\begin{lstlisting}
data Exp : L where
  done : (*@$\uparrow$@*)(Atom (*@$\lto$@*) Exp)
  add : (*@$\uparrow$@*)(Atom (*@$\lto$@*) '+' (*@$\lto$@*) Exp (*@$\lto$@*) Exp)
data Atom : L where
  num : (*@$\uparrow$@*)(NUM (*@$\lto$@*) Atom)
  parens : (*@$\uparrow$@*)('(' (*@$\lto$@*) Exp (*@$\lto$@*) ')' (*@$\lto$@*) Atom)

data O : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
  left : (*@$\uparrow$@*)(&[ n ] &[ b ] '(' (*@$\lto$@*) O (n + 1) b (*@$\lto$@*) O n b)
  num : (*@$\uparrow$@*)(&[ n ] &[ b ] NUM (*@$\lto$@*) D n b (*@$\lto$@*) O n b)
  unexpected : (*@$\uparrow$@*)(&[ n ] (NotStartsWithLP (*@$\lto$@*) O n false)
data D : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
  lookAheadRP : (*@$\uparrow$@*)(&[ n ] &[ b ] ((')' (*@$\otimes$@*) (*@$\top$@*)) (*@$\&$@*) C n b) (*@$\lto$@*) D n b)
  lookAheadNot : (*@$\uparrow$@*)(&[ n ] &[ b ] (NotStartsWithRP (*@$\&$@*) A n b) (*@$\lto$@*) D n b)
data C : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
  closeGood : (*@$\uparrow$@*)(&[ n ] &[ b ] ')' (*@$\lto$@*) D n b (*@$\lto$@*) C (n + 1) b)
  closeBad : (*@$\uparrow$@*)(')' (*@$\lto$@*) C 0 false)
  unexpected : (*@$\uparrow$@*)(&[ n ] NotStartsWithRP (*@$\lto$@*) C n false)
data A : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
  doneGood : (*@$\uparrow$@*)(A 0 true)
  doneBad : (*@$\uparrow$@*)(&[ n ] A (n + 1) false)
  add : (*@$\uparrow$@*)(&[ n ] &[ b ] '+' (*@$\lto$@*) O n b (*@$\lto$@*) A n b)
  unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) ')'(*@$\oplus$@*) NUM) (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) A n false)
\end{lstlisting}
\caption{Associative arithmetic expressions and a corresponding lookahead automaton}
\label{fig:binop-inductive}
\end{figure}
% \subsection{LL(1)}
% %% 4. LL(1) parser

@@

The remaining details of the interpretation of linear terms as parse
transformers is a relatively straightforward extension of existing semantics for
linear logic in monoidal categories \cite{seely89}. In
\ifappendix{\cref{sec:denotational}}\else{the appendix}\fi, we include the denotations
of linear terms and prove that our semantics respects the equational theory of
\theoryabbv --- as well as \cref{ax:dist,ax:disjointness,ax:string-top}.

@@

\newpage
\ifappendix
\appendix
\section{Syntax}
\label{sec:syntax}

>>>>

% -*- fill-column: 80; -*-
\newif\ifarxiv
\arxivtrue

\ifarxiv
\documentclass[acmsmall,nonacm]{acmart}
\else
\documentclass[acmsmall]{acmart}
\fi

\settopmatter{}

@@

%% Keywords
%% comma separated list
\ifarxiv
\else
\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
\fi

@@

This paper begins in \cref{sec:type-theory-examples} by studying small
example programs from \theoryabbv to build intuition.  From there, in
\cref{sec:tt} we provide the syntax, typing, and equational theory of
\theoryname. In \cref{sec:applications} we demonstrate the
applicability of \theoryabbv for relating familiar grammar and
automata formalisms as well as building concrete parsers.  Then in
\cref{sec:semantics-and-metatheory}, we give a denotational semantics
that makes precise the connection between \theoryabbv syntax and
formal grammars. Finally, in \Cref{sec:discussion} we discuss related
and future work.

\ifarxiv\else {\textbf{Extended version:} An extended version of this paper is available in \citet{extendedLambekD}.}\fi

@@

We present a selection of the non-linear type constructors in
\Cref{fig:non-linear-types} and provide the rest in
\ifarxiv{\cref{fig:full-non-linear-types,fig:jdg-eq-nonlinear} of
    \cref{sec:syntax}}\else{the appendix}\fi.  First,
we include universe types $U$ of small non-linear types and $L$ of linear
types. These are defined as
universes ``ala Coquand'' in that we define judgments saying when
non-linear and linear types are \emph{small} and define the universes to internalize
precisely this judgment \cite{coquandPresheafModel,lmcs:7713}. The definition of
smallness is simply that all types are small as long as their sub-formulae are,
with the exception of the two universe types themselves. A formal
description of smallness is given in
\ifarxiv{\cref{fig:small-nonlin-ty,fig:small-lin-ty} of
    \cref{sec:syntax}}\else{the appendix}\fi. These universe
types are needed so that we can define non-linear and linear types by recursion on natural
numbers.
Next, we include standard $\Sigma,\Pi$, empty, unit, Boolean, and natural number
types. More complex inductive types --- such as sum types, list types, and
$Fin~n$ --- can be defined in terms of these primitives, and we give their
encoding in \ifarxiv{\cref{sec:syntax}}\else{the appendix}\fi.

@@

We give the rules for linear type
formation in \cref{fig:linear-formation} and the definition of linear terms in
\cref{fig:linear-terms}. The equational theory for these
types is straightforward $\beta\eta$ equivalence and is included in
\ifarxiv{\cref{fig:jdgeq} of \cref{sec:denotational}}\else{the appendix}\fi.

@@

Next, we introduce the most complex and important linear type
constructors of our development, \emph{indexed inductive linear
types}. We encode these by adding a mechanism for constructing initial
algebras of strictly positive functorial type expressions, following
prior work on inductive types
\cite{nakov_quantitative_2022,altenkirch_indexed_2015}. The syntax is
given in \Cref{fig:iilt}. First, we add a non-linear type $\SPF\,X$ of
\emph{strictly positive functorial} linear type expressions indexed by
a non-linear type $X$. We think of the elements of this type as
syntactic descriptions of linear types that are parameterized by
$X$-many variables standing for linear types that are only used in
strictly positive positions. Accordingly, the $\SPF\,X$ type supports
an operation $\el$ that interprets it as such a type constructor, as
well as an operator $\map$ that defines a functorial action on parse
transformers. The $\SPF\,X$ type supports constructors for a reference
$\Var~x$ to one of the linear type variables, a constant expression
that doesn't mention any type variables $K$, as well as tensor
products and additive conjunction and disjunction of type expressions.
Further, we add equations in \ifarxiv{\cref{fig:spf-act} of \cref{sec:syntax}
  }\else{the appendix }\fi that say that the
$\el$/$\map$ operations correspond to these descriptions of the
constructors.

@@

\begin{figure}
\begin{lstlisting}
data Exp : L where
  done : (*@$\uparrow$@*)(Atom (*@$\lto$@*) Exp)
  add : (*@$\uparrow$@*)(Atom (*@$\lto$@*) '+' (*@$\lto$@*) Exp (*@$\lto$@*) Exp)
data Atom : L where
  num : (*@$\uparrow$@*)('NUM' (*@$\lto$@*) Atom)
  parens : (*@$\uparrow$@*)('(' (*@$\lto$@*) Exp (*@$\lto$@*) ')' (*@$\lto$@*) Atom)

data O : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
  left : (*@$\uparrow$@*)(&[ n ] &[ b ] '(' (*@$\lto$@*) O (n + 1) b (*@$\lto$@*) O n b)
  num : (*@$\uparrow$@*)(&[ n ] &[ b ] 'NUM' (*@$\lto$@*) D n b (*@$\lto$@*) O n b)
  unexpected : (*@$\uparrow$@*)(&[ n ] (NotStartsWithLP (*@$\lto$@*) O n false)
data D : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
  lookAheadRP : (*@$\uparrow$@*)(&[ n ] &[ b ] ((')' (*@$\otimes$@*) (*@$\top$@*)) (*@$\&$@*) C n b) (*@$\lto$@*) D n b)
  lookAheadNot : (*@$\uparrow$@*)(&[ n ] &[ b ] (NotStartsWithRP (*@$\&$@*) A n b) (*@$\lto$@*) D n b)
data C : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
  closeGood : (*@$\uparrow$@*)(&[ n ] &[ b ] ')' (*@$\lto$@*) D n b (*@$\lto$@*) C (n + 1) b)
  closeBad : (*@$\uparrow$@*)(')' (*@$\lto$@*) C 0 false)
  unexpected : (*@$\uparrow$@*)(&[ n ] NotStartsWithRP (*@$\lto$@*) C n false)
data A : Nat (*@$\to$@*) Bool (*@$\to$@*) L where
  doneGood : (*@$\uparrow$@*)(A 0 true)
  doneBad : (*@$\uparrow$@*)(&[ n ] A (n + 1) false)
  add : (*@$\uparrow$@*)(&[ n ] &[ b ] '+' (*@$\lto$@*) O n b (*@$\lto$@*) A n b)
  unexpected : (*@$\uparrow$@*)(&[ n ] ('(' (*@$\oplus$@*) ')'(*@$\oplus$@*) 'NUM') (*@$\lto$@*) (*@$\top$@*) (*@$\lto$@*) A n false)
\end{lstlisting}
\caption{Associative arithmetic expressions and a corresponding lookahead automaton}
\label{fig:binop-inductive}
\end{figure}
% \subsection{LL(1)}
% %% 4. LL(1) parser

@@

The remaining details of the interpretation of linear terms as parse
transformers is a relatively straightforward extension of existing semantics for
linear logic in monoidal categories \cite{seely89}. In
\ifarxiv{\cref{sec:denotational}}\else{the appendix}\fi, we include the denotations
of linear terms and prove that our semantics respects the equational theory of
\theoryabbv --- as well as \cref{ax:dist,ax:disjointness,ax:string-top}.

@@

\newpage
\ifarxiv
\appendix
\section{Syntax}
\label{sec:syntax}
